---
title: "project_checkin"
author: "Tongxin Zhu"
date: "4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(quiet = TRUE)
library(readr)
library(tidyverse)
library(broom)
library(tidyr)
```

# import dataset
```{r}
parenting <- read_csv("C:/Users/74060/Desktop/final_project/parenting.csv", 
                      col_types = cols(CHILDID = col_number(), 
                      P2CHDBTH = col_number(), P2CHLIKE = col_number(), 
                      P2EXPRES = col_number(), P2FLANGR = col_number(), 
                      P2PTHARD = col_number(), P2SACRFC = col_number(), 
                      P2SHOWLV = col_number(), P2TVRUL2 = col_number(), 
                      P2TVRUL3 = col_number(), P2TVRULE = col_number(), 
                      P2WARMCL = col_number(), X12PAR1ED_I = col_character(), 
                      X1PAR1EMP = col_character(), X2INCCAT_I = col_number(), 
                      X_CHSEX_R = col_number(), X_RACETH_R = col_character()))
View(parenting)
```

# select variables of interests: indicators of parenting practices
```{r}
parenting_style <- parenting %>%
  select(P2WARMCL,P2CHLIKE,P2SHOWLV,P2EXPRES,P2PTHARD,P2CHDBTH,P2SACRFC,P2FLANGR)
```

# imputate missing value with column means
```{r}
parenting_style_imp <- replace_na(parenting_style, as.list(colMeans(parenting_style, na.rm = TRUE)))
```

# conduct principle component analysis to make the values of the variables comparable to one another, since the original variables are categorical variables that reflecting the frequency of parenting practices
```{r}
parenting_pca <- prcomp(parenting_style_imp)

pcs_parenting <- parenting_pca %>%
  .$x %>%
  as_tibble()
```

# view the amount of variance explained by each principle component
```{r}
summary(parenting_pca)
```

# plot the distribution of first and second pc
```{r}
as_tibble(parenting_pca$x) %>%
  ggplot(aes(PC1, PC2))+
  geom_point()+
  theme_minimal()
```


? questions: is it because we can only plot two principle components so we pick the two that explain most variance? Also, I don't know how to interpret the plot - seems like the two pcs are negatively correlated? But what does that tell me?

# predict four clusters
```{r}
set.seed(20200426)
parenting_style_kmeans <- kmeans(pcs_parenting,
                                 centers = 4)

tidy(parenting_style_kmeans) %>%
  knitr::kable(digits = 2)
```

# visualize clusters with a scatter plot matrix
```{r}
library(GGally)
bind_cols(parenting_style_imp,
          cluster = parenting_style_kmeans$cluster) %>%
  ggpairs(columns = 1:8,
          mapping = aes(color = factor(cluster)),
          diag = list(continuous = wrap("densityDiag",
                                        alpha =0.8,
                                        color=NA)))

```

? Again, I'm not sure how to interpret the results. Seems like it shows some differences in patterns acroos the four clusters. But how can I take a further analytical step? Any feedbacks and resources are highly appreciated!